The Precambrian paleogeography of Laurentia

### 2021-06-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
updates prior to proofs

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave\endnote{The term Slave Province is rather jarring when read for the first time given that it brings to mind evil human oppression. The geologic Slave Province as well as the eponymous Great Slave Lake get their name from a name given to the First Nation indigenous peoples of the Dene Group who are indigenous to the region. The origin of the name Slave is commonly explained as a French translation of the name given to the Dene people by the Cree people \citep{Britannica-Slave2017a}. Given its negative connotations, the name is not preferred now and there are efforts to remove it from Canadian place names \citep{Mandeville2016a}. Given that this name for the Archean geologic province is deeply entrenched in the geologic literature it will be used here.}+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. Throughout this chapter, I will utilize the term ``accretionary orogenesis'' in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia. ``Collisional orogenesis'' will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure. This usage follows \cite{Staal2020a} who discuss the ambiguity in such a division as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale. Nevertheless, this categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.

@@

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past 1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) magmatism is also shown. The interpreted ages of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic (Table 2) and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig3_Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland). Portions of the apparent polar wander path (APWP) are referred to by the names Logan loop, Keweenawan track, and Grenville loop in the literature and those are labeled next to the associated poles. Upper-right panel:  Paleomagnetic poles from 1800 to 580 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates). The youngest poles from the Ediacaran Period have unusually variable positions as discussed in the text. Lower-left panel: Running mean APWP calculated with a 20 million year moving window. Lower-right panel: Poles for the Archean provinces of Laurentia prior to Laurentia's Paleoproterozoic amalgamation.}
\label{fig:Laurentia_poles}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=5.25 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 20 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig6_Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position. The colors follow Figure \ref{fig:Laurentia_map} where the light grey represents Archean provinces, dark grey represents Paleoproterozoic collisional orogens, and pink/orange represents Paleoproterozoic/Mesoproterozoic accretionary orogens and granitoid intrusions.}
\label{fig:Laurentia_reconstructions}
\end{figure*}

>>>>

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave\endnote{The term Slave Province is rather jarring when read for the first time given that it brings to mind evil human oppression. The geologic Slave Province as well as the eponymous Great Slave Lake get their name from a name given to the First Nation indigenous peoples of the Dene Group who are indigenous to the region. The origin of the name Slave is commonly explained as a French translation of the name given to the Dene people by the Cree people \citep{Britannica-Slave2017a}. Given its negative connotations, the name is not preferred now and there are efforts to remove it from Canadian place names \citep{Mandeville2016a}. Given that this name for the Archean geologic province is deeply entrenched in the geologic literature, it will be used here.}+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. Throughout this chapter, I will utilize the term ``accretionary orogenesis'' in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia. ``Collisional orogenesis'' will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure. This usage follows \cite{Staal2020a} who discuss the ambiguity in such a division as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale. Nevertheless, this categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.

@@

\begin{figure*}
\centering
\includegraphics[width=5.25 in]{../Figures/Fig2_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 20 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig3_Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past 1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) magmatism is also shown. The interpreted ages of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic (Table 2) and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

@@

\begin{figure*}[!h]
\centering
\includegraphics[width=\textwidth]{../Figures/Fig4_Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland). Portions of the apparent polar wander path (APWP) are referred to by the names Logan loop, Keweenawan track, and Grenville loop in the literature and those are labeled next to the associated poles. Upper-right panel:  Paleomagnetic poles from 1800 to 580 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates). The youngest poles from the Ediacaran Period have unusually variable positions as discussed in the text. Lower-left panel: Running mean APWP calculated with a 20 million year moving window. Lower-right panel: Poles for the Archean provinces of Laurentia prior to Laurentia's Paleoproterozoic amalgamation.}
\label{fig:Laurentia_poles}
\end{figure*}

@@

@@

\begin{figure*}[!h]
\centering
\includegraphics[width=\textwidth]{../Figures/Fig6_Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position. The colors follow Figure \ref{fig:Laurentia_map} where the light grey represents Archean provinces, dark grey represents Paleoproterozoic collisional orogens, and pink/orange represents Paleoproterozoic/Mesoproterozoic accretionary orogens and granitoid intrusions.}
\label{fig:Laurentia_reconstructions}
\end{figure*}
